---
title: "SIMULATING BIODIVERSITY PATTERNS: NEUTRAL, COMPETITIVE & INVASION MODELS"
author: "Phoebe Jekielek and Sarah Risley"
date: "Fall, 2022"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(roleLite)
library(vegan)
library(pika)
library(tidyverse)
library(knitr)
```

**Introduction**        

Hubbell's neutral theory of biodiversity and biogeography highlights the importance of unifying ecological processes--birth, death, and immigration--and offers a null model that can provide insight into current and future species-abundance distributions. Neutral theory is commonly used in ecology to describe and understand biodiversity and evolution patterns, and although its applications are expanding, it primarily supports theoretical exploration (Rosindell et al., 2011). Although neutral theory had been applied in the marine context (see Alzate et al., 2019, Dolan et al., 2007, Wootton, 2005, among others), it is not commonly used in marine ecosystems or for species that are targeted by marine fisheries. 

Biodiversity indices have been developed over the years to help us compare and measure biodiversity across scales, taxa, trophic levels or regions. Species richness is perhaps the simplest of the diversity metrics, representing the total number of species represented in a population without taking species abundance or relative abundance into consideration (Whittaker 1972). Both the Shannon-Weiner index (Shannon & Weaver 1949) and the Simpson's Index take into account species richness and abundance distribution of a species. The Shannon-Weiner index measures the amount of uncertainty of picking the same species twice from a random draw from all of the species in a community. Therefore, a high level of uncertainty indicates a higher species diversity. The Simpson's Index measures probability that two individuals randomly sampled belong to different species. Again, the higher the Simpson index, the higher the diversity. 

In this study, we seek to explore how neutral theory can be used to understand biodiversity patterns in intertidal marine ecosystems, with a focus on commercially important shellfish species. Using classical biodiversity indices, will compare the neutral theory model to a competitive model and assess how well each model represents biodiversity patterns in a marine ecosystem. Specifically, we will model the biodiversity patterns that occur under immigration pulses from rare species. This approach aims to model recent species introductions observed in many of Maine's intertidal marine ecosystems, namely the introduction of wild American oysters and increases in invasive green crab populations. This study will inform future analyses and help to identify potential forces driving biodiversity patterns in these systems.    


**Study System**        

The Damariscotta River estuary, located in midcoast Maine, hosts a diversity of coastal marine habitats ranging from rocky shores to soft sediment intertidal areas and eelgrass beds. The estuary is home to extensive human activity, including recreational boating, commercial fishing, and shellfish aquaculture. The Damariscotta has more than a dozen marine aquaculture operations that total 0.93 km², where farmers grow American oysters, mussels, scallops, and kelp (Maine DMR, 2022). The biodiversity in the Damariscotta is similar to other estuaries in the Gulf of Maine and varies along habitat type and physical and biological gradients (Britsch, 2021). 

The intertidal mudflat is the focus of this study. This habitat hosts a diversity of shellfish, marine worm, and arthropod species. Of commercial interest are soft-shell clams (*Mya arenara*), quahogs (*Mercenaria mercenaria*), razor clams (*Ensis directus*), blood worms (*Glycera dibranchiata*), and sand worms (*Nereis virens*). Within the past 20 years commercial shellfish harvesters have observed the appearance of wild American oyster (*Crassostrea virginica*), a previously extirpated species now reappearing ostensibly as a result of aquaculture activity (Delago, 2021; Risley, 2022). Green crabs (*Carcinus maenas*), an invasive species originally from the Mediterranean, have also experienced a recent growth in population abundance, likely linked to warming ocean temperatures. Green crabs prey on shellfish species, particularly the delicate soft-shell clam, and have resulted in declines in clam populations as their number grow (Tan and Beal, 2015). 

Therefore, the Damariscotta has two 'rare' species whose growth and expansion could affect the biodiversity of the whole system. This study seeks two explore how increases in rare species, like the wild American oyster and the green crab, can influence biodiversity within the ecosystem. 


**Methods**         

For this project we will use R Studio to simulate both a neutral theory model and a competitive model for the system. We utilized the roleLite R package (xversion 0.1.0, Rominger) and then utilized the untbSim function to run a neutral simulation for a null model evaluation with the following parameters: J = 1000 (number of individuals in local community), Sm = 50 (number of species in meta community), Jm = 10000 (number of individuals in meta community), nu = 0 (speciation probability), m = 0.1 (immigration probability), and niter = 10000 (number of iterations to run). These parameters were applied across all models. 

Additionally, we ran a competitive simulation model using the compSim function. This simulation used a competitive matrix with intraspecies competition removed in order to increase average niche differentiation. Both models were visualized using the final time step of a species abundance distribution (SAD) and biodiversity indices (Simpsons, Shannons, Richness) were calculated for each model. 

An invasion model was generated to simulate the amplification of a rare species. This model consisted of a pre-invasion simulation merged with a post-invasion simulation in order to evaluate the effects of this amplification. We visualized SAD's for the invasion model at three timesteps -10,000, 13,000, and 20,000 iterations - in order to visualize biodiersity patterns immediately before an invasion (through 10,000 iterations), shortly after the invasion (from 10,000 to 13,000 iterations), and long after the invasion (up to 20,000 iterations). Biodiversity indices were calculated for each of these timesteps, respectively. The metacommunity was modified in the post-simulation to amplify the rare species by a factor of two (the rare species were twice as common). Both pre- and post-simulations used the untbSim package with the same parameters used in the previously-mentioned neutral and competitive models. 

Finally, measures for Simpsons and Shannons Diversity Indices and Species Richness, were calculated to compare and contrast neutral, competitive and merged models for resulting biodiversity impacts. 



**Results** (PJ)                


```{r, include=FALSE}
###NEUTRAL SIMULATION###
##Run simulation
neut <- untbSim(J = 1000, # number of individuals in local community
                Sm = 50, # number of species in meta community
                Jm = 10000, # number of individuals in meta community
                nu = 0, # speciation probability
                m = 0.1, # immigration probability 
                niter = 10000) # number of iteration to run

##Passing to getAbund
neutAbund <- getAbund(neut)
#log-series in meta community, initialize with random pull form meta comm for start of local comm

# returns a data.frame, see `?getAbund` for more detials
head(neutAbund)

###COMPETITIVE SIMULATION###

S <- 50 # number of species
#best practice to establish variable outside of fxn call 

# make a competition matrix
a <- matrix(runif(S^2, 0, 5), nrow = S) 
diag(a) <- runif(S, 1, 10) #increasing avg niche differentiation 

# run sim
comp <- compSim(J = 1000, # number of individuals in local community
                Sm = S, # number of species in meta community
                Jm = 10000, # number of individuals in meta community
                nu = 0, # speciation probability
                m = 0.1, # immigration probability
                alpha = a, # competition matrix
                niter = 10000) # number of iteration to run

compAbund <- getAbund(comp)
head(compAbund)
```


*Neutral Model SADs*                


```{r, echo = FALSE}
##Looking at the final SAD for neut
finalSADneut <- neutAbund$abund[neutAbund$tstep == max(neutAbund$tstep)]

plot(sort(finalSADneut, decreasing = TRUE), log = 'y', 
     xlab = 'Rank', ylab = 'Abundance',
     main = 'Neutral SAD')
```

<<<<<<< HEAD
Resulting species abundance from the neutral model (277) was higher than that for the competitive model (262). 
-Max abund = 277
=======
>>>>>>> 78d3fe991645c3f135d6ddc6c771b3ddec458b9d



*Competitive Model SADs*                  

```{r, echo = FALSE}
##Looking at the final SAD for comp
finalSADComp <- compAbund$abund[compAbund$tstep == max(compAbund$tstep)]

plot(sort(finalSADComp, decreasing = TRUE), log = 'y', 
     xlab = 'Rank', ylab = 'Abundance',
     main = 'Competitive SAD')
```
<<<<<<< HEAD

-Max abund = 262
-More species with lower abundances 
-Overall, graphs look similar 
-Linear slope 
=======
Resulting species abundance from the neutral model (277) was higher than that for the competitive model (262). Overall, the rank abundance plots for the neutral and competitive models are very similar with the exception of the competitive model having more species with lower abundances. 
>>>>>>> 78d3fe991645c3f135d6ddc6c771b3ddec458b9d



*Invasion Model SADs*             

```{r, include=FALSE}
###PURTUBATIONS###

##Neutral: rare species becomes abundant
# parameter set-up
S <- 50 
J <- 1000
nu <- 0
m <- 0.1
niter <- 10000

# make meta SAD outside of `untbSim` so we can manipulate it to reflect 
# invasion...hang tight, you'll see
metaSAD <- pika::rlseries(S, 0.001) 

# run sim pre-invasion
untbPre <- untbSim(J = J, nu = nu, m = m, niter = niter, initMeta = metaSAD)

#don't need to specify Sm and Jm 

##Modifying metacommunity so least common species is 2x's as common
newMetaSAD <- metaSAD
irare <- which.min(newMetaSAD)[1] # index of a rare species
newMetaSAD[irare] <- 2 * max(newMetaSAD)

# now we can pass this new meta SAD to the sim function, and pass the final 
# state of the pre-invasion simulation as the initial condition for this sim:
untbPost <- untbSim(J = J, nu = nu, m = m, niter = niter, 
                    initLocal = untbPre[nrow(untbPre), ], 
                    initMeta = newMetaSAD)

##Combining the two simulations
abundPre <- getAbund(untbPre)
abundPost <- getAbund(untbPost)

# note: first timestep of post is last of pre, so we can discard one
abundPost <- abundPost[abundPost$tstep != 0, ]

# update timestep values of post to come *after* pre
abundPost$tstep <- abundPost$tstep + max(abundPre$tstep)

# combine
abundAll <- rbind(abundPre, abundPost)

# visualize abundance timeseries for invader
abundInvader <- abundAll[abundAll$spID == irare, ]

plot(abundInvader$tstep, abundInvader$abund, type = 'b',
     xlab = 'Timestep', ylab = 'Abundance')

```


The results from the invasion simulation varied. The pre-invasion and shortly-after invasion results are very similar to each other with steep drops from max abundances across all time steps. For example, max abundance of species pre-invasion is 431, post-short-term invasion is 409 and post-long-term invasion is 354. 


```{r, echo = FALSE}
#Right before invasion
plot(sort(abundPre$abund[abundPre$tstep == max(abundPre$tstep)], 
          decreasing = TRUE), 
     log = 'y', 
     xlab = 'Rank', ylab = 'Abundance', main = 'Right before invasion')

#Shortly after invasion
plot(sort(abundPost$abund[abundPost$tstep == 13000], 
          decreasing = TRUE), 
     log = 'y', 
     xlab = 'Rank', ylab = 'Abundance', main = 'Shortly after invasion')

#Long after invasion
plot(sort(abundPost$abund[abundPost$tstep == max(abundPost$tstep)], 
          decreasing = TRUE), 
     log = 'y', 
     xlab = 'Rank', ylab = 'Abundance', main = 'Long after invasion')
```
 

*Calculating Hill Numbers: Model Biodiversity Indices*                

```{r, include = FALSE}
###HILL NUMBERS###
# params set-up
Sm <- 50
sadMeta <- rlseries(Sm, 0.001)
J <- 1000
nu <- 0
m <- 0.1
niter <- 10000

## Neutral sim hill 
hillz <- replicate(500, # number of times to replicate
                   { # put all the code you need to replicate inside the { }
                     oneNeut <- untbSim(J = J, nu = nu, m = m, niter = niter, 
                                        initMeta = sadMeta)
                     
                     oneFinalSAD <- getAbund(oneNeut, tstep = nrow(oneNeut))
                     
                     # hill numbers
                     theseHill <- renyi(oneFinalSAD$abund, scales = 0:2, 
                                        hill = TRUE)
                     
                     # last line inside the {} should be the output you want
                     return(theseHill)
                   })

# Summarizing this raw simulation, mean and SD are good options
neut_hillMeans <- apply(hillz, 1, mean)
neut_hillSDs <- apply(hillz, 1, sd)

## Comp sim hill 
hillz_2 <- replicate(500, # number of times to replicate
                   { # put all the code you need to replicate inside the { }
                           oneComp <- compSim(J = J, Sm = Sm, Jm = 10000, 
                                              nu = nu,m = m, alpha = a, niter = niter) 
                           
                           oneFinalSAD <- getAbund(oneComp, tstep = nrow(oneComp))
                           
                           # hill numbers
                           theseHill <- renyi(oneFinalSAD$abund, scales = 0:2, 
                                              hill = TRUE)
                           
                           # last line inside the {} should be the output you want
                           return(theseHill)
                   })


# Summarizing this raw simulation, mean and SD are good options
comp_hillMeans <- apply(hillz_2, 1, mean)
comp_hillSDs <- apply(hillz_2, 1, sd)


##Invasion sims

# Pre
hillz_4 <- replicate(500, # number of times to replicate
                     { # put all the code you need to replicate inside the { }
                             metaSAD <- pika::rlseries(S, 0.001) 
                             untbPre <- untbSim(J = J, nu = nu, m = m, niter = niter,
                                                initMeta = metaSAD) # pre invasion
                             
                             newMetaSAD <- metaSAD # modifying meta community
                             irare <- which.min(newMetaSAD)[1] # index of a rare species
                             newMetaSAD[irare] <- 2 * max(newMetaSAD) # amplifying rare x2
                             
                             untbPost <- untbSim(J = J, nu = nu, m = m, niter = niter, 
                                                 initLocal = untbPre[nrow(untbPre), ], 
                                                 initMeta = newMetaSAD) # post invasion
                             
                             abundPre <- getAbund(untbPre) # combining sims
                             abundPost <- getAbund(untbPost)
                             abundPost <- abundPost[abundPost$tstep != 0, ] # removing overlap
                             abundPost$tstep <- abundPost$tstep + max(abundPre$tstep) # timesteps post after pre
                             abundAll <- rbind(abundPre, abundPost) # combining
                             
                             oneFinalSAD <- abundPre$abund[abundPre$tstep == max(abundPre$tstep)]
                             
                             
                             # hill numbers
                             theseHill <- renyi(oneFinalSAD, scales = 0:2, 
                                                hill = TRUE)
                             
                             # last line inside the {} should be the output you want
                             return(theseHill)
                             
                     })

# Summarizing this raw simulation, mean and SD are good options
preinvas_hillMeans <- apply(hillz_4, 1, mean)
preinvas_hillSDs <- apply(hillz_4, 1, sd)

# Short-term post
hillz_5 <- replicate(500, # number of times to replicate
                     { # put all the code you need to replicate inside the { }
                             metaSAD <- pika::rlseries(S, 0.001) 
                             untbPre <- untbSim(J = J, nu = nu, m = m, niter = niter,
                                                initMeta = metaSAD) # pre invasion
                             
                             newMetaSAD <- metaSAD # modifying meta community
                             irare <- which.min(newMetaSAD)[1] # index of a rare species
                             newMetaSAD[irare] <- 2 * max(newMetaSAD) # amplifying rare x2
                             
                             untbPost <- untbSim(J = J, nu = nu, m = m, niter = niter, 
                                                 initLocal = untbPre[nrow(untbPre), ], 
                                                 initMeta = newMetaSAD) # post invasion
                             
                             abundPre <- getAbund(untbPre) # combining sims
                             abundPost <- getAbund(untbPost)
                             abundPost <- abundPost[abundPost$tstep != 0, ] # removing overlap
                             abundPost$tstep <- abundPost$tstep + max(abundPre$tstep) # timesteps post after pre
                             abundAll <- rbind(abundPre, abundPost) # combining
                             
                             oneFinalSAD <- abundPost$abund[abundPost$tstep == 13000]
                             
                             
                             # hill numbers
                             theseHill <- renyi(oneFinalSAD, scales = 0:2, 
                                                hill = TRUE)
                             
                             # last line inside the {} should be the output you want
                             return(theseHill)
                             
                     })
# Summarizing this raw simulation, mean and SD are good options
postinvas_hillMeans <- apply(hillz_5, 1, mean)
postinvas_hillSDs <- apply(hillz_5, 1, sd)

# Long-Term Post
hillz_6 <- replicate(500, # number of times to replicate
                     { # put all the code you need to replicate inside the { }
                             metaSAD <- pika::rlseries(S, 0.001) 
                             untbPre <- untbSim(J = J, nu = nu, m = m, niter = niter,
                                                initMeta = metaSAD) # pre invasion
                             
                             newMetaSAD <- metaSAD # modifying meta community
                             irare <- which.min(newMetaSAD)[1] # index of a rare species
                             newMetaSAD[irare] <- 2 * max(newMetaSAD) # amplifying rare x2
                             
                             untbPost <- untbSim(J = J, nu = nu, m = m, niter = niter, 
                                                 initLocal = untbPre[nrow(untbPre), ], 
                                                 initMeta = newMetaSAD) # post invasion
                             
                             abundPre <- getAbund(untbPre) # combining sims
                             abundPost <- getAbund(untbPost)
                             abundPost <- abundPost[abundPost$tstep != 0, ] # removing overlap
                             abundPost$tstep <- abundPost$tstep + max(abundPre$tstep) # timesteps post after pre
                             abundAll <- rbind(abundPre, abundPost) # combining
                             
                             oneFinalSAD <- abundPost$abund[abundPost$tstep == max(abundPost$tstep)]
                             
                             
                             # hill numbers
                             theseHill <- renyi(oneFinalSAD, scales = 0:2, 
                                                hill = TRUE)
                             
                             # last line inside the {} should be the output you want
                             return(theseHill)
                             
                     })

# Summarizing this raw simulation, mean and SD are good options
longinvas_hillMeans <- apply(hillz_6, 1, mean)
longinvas_hillSDs <- apply(hillz_6, 1, sd)

```


The competitive model maintains the highest species richness, Shannons Index, and Simpsons Index values followed by the neutral, post-invasion and pre-invasion simulations, respectively. 

- Simpson's: relative abundances, how evenly the abundances are distributed
  -Greatest diversity of species in competitive model
  -Followed by Neutral
  -Long term after invasion, then post invasion, then pre-invasion is lowest
  -Relatively invasion model is much lower than both neutral and competitive 
  

-Species richness: count of species in community 
  -Neutral model has greatest richness
  -Competitive and pre-invasion are the same
  -Richness declines post invasion, with lowest at long term after invasion time step 
  -So fewer species in post-invasion model, but greater evenness regarding abundances across species 

-Shannon's: species richness
  -Neutral and competitive similar and most diverse
  -Invasion model on the whole is lower
  -Lowest pre, then after invasion, with short-term post-invasion having the highest of the three 

the importance of rare species decreases in order species richness > Shannon index > Simpson index        



```{r, echo = FALSE, results='asis'}
## Compiling
hills_all <- rbind(neut_hillMeans,neut_hillSDs,
             comp_hillMeans, comp_hillSDs,
             preinvas_hillMeans, preinvas_hillSDs,
             postinvas_hillMeans, postinvas_hillSDs,
             longinvas_hillMeans,longinvas_hillSDs)

hills_all <-round(hills_all, digits = 2)

hills_all <- as.data.frame(hills_all)

hills_all <- hills_all %>% 
        rename( "Species Richness" = "0",
                "Shannons Index" = "1",
                "Inverse Simpsons" = "2")

row.names(hills_all) <- c("Neutral Model","SD Neut","Competitive Model","SD Comp","Pre-Invasion Model", "SD Invas1",
              "Short-Term Post Invasion","SD Invas2","Long-Term Post Invasion","SD Invas3")

kable(hills_all, caption = "Mean Biodiversity Indices")
```



**Discussion**

-Neutral and competitive are very similar, give similar predictions
-Competitive model is more diverse 
-Niche differentiation increases in competitive model, therefore more likely to be diverse than neutral


- (He&Hu 2005): Like any other diversity indices, the Simpson index is also sensitive to sample size (increase with sample size) although its overall performance against several criteria is superior to many other indices such as the Shannon information entropy . The relationship between the NTB and Simpson diversity suggests that diversity patterns of the metacommunity can also be equally described by the Simpson index. This relationship provides an alternative approach to interpret and estimate the fundamental biodiversity parameter for the metacommunity.


- Levine&HRL: Our results support the hypothesis that niche differences strongly stabilize coexistence. Most importantly, our findings provide strong empirical support for the critical role niche differences have in stabilizing species diversity, one of the longest unresolved problems in ecology.In an experiment manipulating species’ relative abundances, population growth rates increased when species became rare—the demographic signature of niche differences. Our work thus provides strong evidence that species differences have a critical role in stabilizing species diversity.





*Things to Consider*

Applying neutral models to marine systems can be slightly more complex than applying the same model to terrestrial ecosystems. For one, dispersal limitation may play a more prominent role in ocean systems where both environmental factors (e.g. ocean currents) and marine life stages (e.g. planktonic larval stages) can influence community structure and biodiversity (Srisvastava and Pavel, 2013).

The invasion model 
- Discussion around special considerations for marine systems (Chust et al., 2016)
-Refutes NTB (Dornela et al., 2006)
  - How does the scale of observation influence model results? (Chisholm and Lichstein, 2009)
  -What about harvesting activity (disturbance)? How does that play into the NTB model?
  - - Maa (2014): Neutral models provide a robust null hypothesis because  they can provide estimates of beta diversity based on the simplest metacommunity scenario. However, neutral models can be used to detect disturbances in two different ways: First, data reject neutral predictions because the real assemblages vary too much or too little in terms of species composition and abundance (Dornelas et al. 2006; Caruso et al. 2012a,b); second, communities are really assembled under neutral dynamics, and disturbance directly affects neutral parameters, for example, by decreasing dispersal via increasing habitat fragmentation (Hubbell 2001) or by affecting some fundamental demographic parameters (Dornelas 2010). In this study, we basically used neutral models in the first sense because we believe that in observational studies, robust conclusions can be obtained only when sound statistical null hypotheses are rejected (Gotelli and Ulrich 2012).
  
  
  -How does genetic manipulation from AQ influence biodiversity? *Maybe think of various impacts of AQ on biodiversity, not just genetic manipulation...
  -- Diana (2009): Many impacts of AQ on biodiversity - introduction of non-natives to local populations, introduction of less-fit genes, introduction of disease, effluents and waste, harvesting of bait fish for aq feed...all of these can have negative impacts on biodiversity. 
  -Connolly et al., 2014
  
*Next Steps*

-What worked and didn't work 




**References**

Alzate, A., Janzen, T., Bonte, D., Rosindell, J., & Etienne, R. S. (2019). A simple spatially explicit neutral model explains the range size distribution of reef fishes. Global Ecology and Biogeography, 28(7), 875–890. https://doi.org/10.1111/geb.12899

Britsch, M. L. (2021). Marine Aquaculture in Maine: Understanding Diverse Perspectives and Interactions at Multiple Scales. University of Maine.

Delago, D. F. (2021). Investigating Larval Spillover From Oyster Aquaculture Through Geospatial Habitat Suitability Index Modeling: A Damariscotta River Estuary Case Study. University of New England.

Diana, J. S. (2009). Aquaculture production and biodiversity conservation. Bioscience, 59(1), 27-38.

Dolan, J. R., Ritchie, M. E., & Ras, J. (2007). The “neutral” community structure of planktonic herbivores, tintinnid ciliates of the microzooplankton, across the SE Tropical Pacific Ocean. Biogeosciences, 4(3), 297–310. https://doi.org/10.5194/bg-4-297-2007

Levine, J. M., & HilleRisLambers, J. (2009). The importance of niches for the maintenance of species diversity. Nature, 461(7261), 254-257.

Maine DMR. (2022). Aquaculture Map: Maine Department of Marine Resources. https://www.maine.gov/dmr/aquaculture/leases/aquaculturemap.html

Risley, S. (2022). Linking Local Knowledge & Community Science in Support of Coastal Marine Stewardship (3564) [The University of Maine]. Electronic Theses and Dissertations. https://digitalcommons.library.umaine.edu/etd/3564

Rosindell, J., Hubbell, S. P., & Etienne, R. S. (2011). The Unified Neutral Theory of Biodiversity and Biogeography at Age Ten. Trends in Ecology & Evolution, 26(7), 340–348. https://doi.org/10.1016/j.tree.2011.03.024

Shannon, C.E. & Weaver, W. (1949) The Mathematical Theory of
Communication. University of Illinois Press, Urbana.

Srivastava, D., & Kratina, P. (2013). Is dispersal limitation more prevalent in the ocean? Oikos, 122, 298–300.

Tan, E. B. P., & Beal, B. F. (2015). Interactions between the invasive European green crab, Carcinus maenas (L.), and juveniles of the soft-shell clam, Mya arenaria L., in eastern Maine, USA. Journal of Experimental Marine Biology and Ecology, 462, 62–73. https://doi.org/10.1016/j.jembe.2014.10.021

Whittaker, R. H. 1972. Evolution and measurement of species diversity. Taxon 21:213–251.

Wootton, J. T. (2005). Field parameterization and experimental test of the neutral theory of biodiversity—Nature. Nature, 433(7023), 309–312. https://doi.org/10.1038/nature03211







