---
title: "SIMULATING BIODIVERSITY PATTERNS: NEUTRAL, COMPETITIVE & INVASION MODELS"
author: "Phoebe Jekielek and Sarah Risley"
date: "Fall, 2022"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(roleLite)
library(vegan)
library(pika)
library(tidyverse)
library(knitr)
```

**Introduction**        

Hubbell's neutral theory of biodiversity and biogeography highlights the importance of unifying ecological processes--birth, death, and immigration--and offers a null model that can provide insight into current and future species-abundance distributions. Neutral theory is commonly used in ecology to describe and understand biodiversity and evolution patterns, and although its applications are expanding, it primarily supports theoretical exploration (Rosindell et al., 2011). Although neutral theory had been applied in the marine context (see Alzate et al., 2019, Dolan et al., 2007, Wootton, 2005, among others), it is not commonly used in marine ecosystems or for species that are targeted by marine fisheries. 

Biodiversity indices have been developed over the years to help us compare and measure biodiversity across scales, taxa, trophic levels or regions. Species richness is perhaps the simplest of the diversity metrics, representing the total number of species represented in a population without taking species abundance or relative abundance into consideration (Whittaker 1972). Both the Shannon-Weiner index (Shannon & Weaver 1949) and the Simpson's Index take into account species richness and abundance distribution of a species. The Shannon-Weiner index measures the amount of uncertainty of picking the same species twice from a random draw from all of the species in a community. Therefore, a high level of uncertainty indicates a higher species diversity. The Simpson's Index measures probability that two individuals randomly sampled belong to different species. Again, the higher the Simpson index, the higher the diversity. 

In this study, we seek to explore how neutral theory can be used to understand biodiversity patterns in intertidal marine ecosystems, with a focus on commercially important shellfish species. Using classical biodiversity indices, will compare the neutral theory model to a competitive model and assess how well each model represents biodiversity patterns in a marine ecosystem. Specifically, we will model the biodiversity patterns that occur under immigration pulses from rare species. This approach aims to model recent species introductions observed in many of Maine's intertidal marine ecosystems, namely the introduction of wild American oysters and increases in invasive green crab populations. This study will inform future analyses and help to identify potential forces driving biodiversity patterns in these systems.    


**Study System**        

The Damariscotta River estuary, located in midcoast Maine, hosts a diversity of coastal marine habitats ranging from rocky shores to soft sediment intertidal areas and eelgrass beds. The estuary is home to extensive human activity, including recreational boating, commercial fishing, and shellfish aquaculture. The Damariscotta has more than a dozen marine aquaculture operations that total 0.93 km², where farmers grow American oysters, mussels, scallops, and kelp (Maine DMR, 2022). The biodiversity in the Damariscotta is similar to other estuaries in the Gulf of Maine and varies along habitat type and physical and biological gradients (Britsch, 2021). 

The intertidal mudflat is the focus of this study. This habitat hosts a diversity of shellfish, marine worm, and arthropod species. Of commercial interest are soft-shell clams (*Mya arenara*), quahogs (*Mercenaria mercenaria*), razor clams (*Ensis directus*), blood worms (*Glycera dibranchiata*), and sand worms (*Nereis virens*). Within the past 20 years commercial shellfish harvesters have observed the appearance of wild American oyster (*Crassostrea virginica*), a previously extirpated species now reappearing ostensibly as a result of aquaculture activity (Delago, 2021; Risley, 2022). Green crabs (*Carcinus maenas*), an invasive species originally from the Mediterranean, have also experienced a recent growth in population abundance, likely linked to warming ocean temperatures. Green crabs prey on shellfish species, particularly the delicate soft-shell clam, and have resulted in declines in clam populations as their number grow (Tan and Beal, 2015). 

Therefore, the Damariscotta has two 'rare' species whose growth and expansion could affect the biodiversity of the whole system. This study seeks two explore how increases in rare species, like the wild American oyster and the green crab, can influence biodiversity within the ecosystem. 


**Methods**         

For this project we will use R Studio to simulate both a neutral theory model and a competitive model for the system. We utilized the roleLite R package (xversion 0.1.0, Rominger) and then utilized the untbSim function to run a neutral simulation for a null model evaluation with the following parameters: J = 1000 (number of individuals in local community), Sm = 50 (number of species in meta community), Jm = 10000 (number of individuals in meta community), nu = 0 (speciation probability), m = 0.1 (immigration probability), and niter = 10000 (number of iterations to run). These parameters were applied across all models. 

Additionally, we ran a competitive simulation model using the compSim function. This simulation used a competitive matrix with intraspecies competition removed in order to increase average niche differentiation. Both models were visualized using the final time step of a species abundance distribution (SAD) and biodiversity indices (Simpsons, Shannons, Richness) were calculated for each model. 

An invasion model was generated to simulate the amplification of a rare species. This model consisted of a pre-invasion simulation merged with a post-invasion simulation in order to evaluate the effects of this amplification. We visualized SAD's for the invasion model at three timesteps -10,000, 13,000, and 20,000 iterations - in order to visualize biodiersity patterns immediately before an invasion (through 10,000 iterations), shortly after the invasion (from 10,000 to 13,000 iterations), and long after the invasion (up to 20,000 iterations). The metacommunity was modified in the post-simulation to amplify the rare species by a factor of two (the rare species were twice as common). Both pre- and post-simulations used the untbSim package with the same parameters used in the previously-mentioned neutral and competitive models. 

In order to calculate Hill numbers with robust statistical power we used the replicate function to run 500 iterations of our simulation. This created 500 distinct species abundance distributions which were then input into the "Renyi" function from the "Vegan" package to return 500 values of the biodiversity indices from which generated means and standard deviations for all models. These values were then compared across neutral, competitive and merged models for resulting biodiversity impacts. 

**Results** (PJ)                


```{r, include=FALSE}
###NEUTRAL SIMULATION###
##Run simulation
neut <- untbSim(J = 1000, # number of individuals in local community
                Sm = 50, # number of species in meta community
                Jm = 10000, # number of individuals in meta community
                nu = 0, # speciation probability
                m = 0.1, # immigration probability 
                niter = 10000) # number of iteration to run

##Passing to getAbund
neutAbund <- getAbund(neut)
#log-series in meta community, initialize with random pull form meta comm for start of local comm

# returns a data.frame, see `?getAbund` for more detials
head(neutAbund)

###COMPETITIVE SIMULATION###

S <- 50 # number of species
#best practice to establish variable outside of fxn call 

# make a competition matrix
a <- matrix(runif(S^2, 0, 5), nrow = S) 
diag(a) <- runif(S, 1, 10) #increasing avg niche differentiation 

# run sim
comp <- compSim(J = 1000, # number of individuals in local community
                Sm = S, # number of species in meta community
                Jm = 10000, # number of individuals in meta community
                nu = 0, # speciation probability
                m = 0.1, # immigration probability
                alpha = a, # competition matrix
                niter = 10000) # number of iteration to run

compAbund <- getAbund(comp)
head(compAbund)
```


*Neutral Model SADs*                


```{r, echo = FALSE}
##Looking at the final SAD for neut
finalSADneut <- neutAbund$abund[neutAbund$tstep == max(neutAbund$tstep)]

plot(sort(finalSADneut, decreasing = TRUE), log = 'y', 
     xlab = 'Rank', ylab = 'Abundance',
     main = 'Neutral SAD')
```

*Competitive Model SADs*                  

```{r, echo = FALSE}
##Looking at the final SAD for comp
finalSADComp <- compAbund$abund[compAbund$tstep == max(compAbund$tstep)]

plot(sort(finalSADComp, decreasing = TRUE), log = 'y', 
     xlab = 'Rank', ylab = 'Abundance',
     main = 'Competitive SAD')
```

*Invasion Model SADs*             

```{r, include=FALSE}
###PURTUBATIONS###

##Neutral: rare species becomes abundant
# parameter set-up
S <- 50 
J <- 1000
nu <- 0
m <- 0.1
niter <- 10000

# make meta SAD outside of `untbSim` so we can manipulate it to reflect 
# invasion...hang tight, you'll see
metaSAD <- pika::rlseries(S, 0.001) 

# run sim pre-invasion
untbPre <- untbSim(J = J, nu = nu, m = m, niter = niter, initMeta = metaSAD)

#don't need to specify Sm and Jm 

##Modifying metacommunity so least common species is 2x's as common
newMetaSAD <- metaSAD
irare <- which.min(newMetaSAD)[1] # index of a rare species
newMetaSAD[irare] <- 2 * max(newMetaSAD)

# now we can pass this new meta SAD to the sim function, and pass the final 
# state of the pre-invasion simulation as the initial condition for this sim:
untbPost <- untbSim(J = J, nu = nu, m = m, niter = niter, 
                    initLocal = untbPre[nrow(untbPre), ], 
                    initMeta = newMetaSAD)

##Combining the two simulations
abundPre <- getAbund(untbPre)
abundPost <- getAbund(untbPost)

# note: first timestep of post is last of pre, so we can discard one
abundPost <- abundPost[abundPost$tstep != 0, ]

# update timestep values of post to come *after* pre
abundPost$tstep <- abundPost$tstep + max(abundPre$tstep)

# combine
abundAll <- rbind(abundPre, abundPost)

# visualize abundance timeseries for invader
abundInvader <- abundAll[abundAll$spID == irare, ]

plot(abundInvader$tstep, abundInvader$abund, type = 'b',
     xlab = 'Timestep', ylab = 'Abundance')

```

The results from the invasion simulation varied. The pre-invasion and shortly-after invasion results are very similar to each other with steep drops from max abundances across all time steps. For example, max abundance of species pre-invasion is 431, post-short-term invasion is 409 and post-long-term invasion is 354. 

```{r, echo = FALSE}
#Right before invasion
plot(sort(abundPre$abund[abundPre$tstep == max(abundPre$tstep)], 
          decreasing = TRUE), 
     log = 'y', 
     xlab = 'Rank', ylab = 'Abundance', main = 'Right before invasion')

#Shortly after invasion
plot(sort(abundPost$abund[abundPost$tstep == 13000], 
          decreasing = TRUE), 
     log = 'y', 
     xlab = 'Rank', ylab = 'Abundance', main = 'Shortly after invasion')

#Long after invasion
plot(sort(abundPost$abund[abundPost$tstep == max(abundPost$tstep)], 
          decreasing = TRUE), 
     log = 'y', 
     xlab = 'Rank', ylab = 'Abundance', main = 'Long after invasion')
```
 

*Calculating Hill Numbers: Model Biodiversity Indices*                

```{r, include = FALSE}
###HILL NUMBERS###
# params set-up
Sm <- 50
sadMeta <- rlseries(Sm, 0.001)
J <- 1000
nu <- 0
m <- 0.1
niter <- 10000

## Neutral sim hill 
hillz <- replicate(500, # number of times to replicate
                   { # put all the code you need to replicate inside the { }
                     oneNeut <- untbSim(J = J, nu = nu, m = m, niter = niter, 
                                        initMeta = sadMeta)
                     
                     oneFinalSAD <- getAbund(oneNeut, tstep = nrow(oneNeut))
                     
                     # hill numbers
                     theseHill <- renyi(oneFinalSAD$abund, scales = 0:2, 
                                        hill = TRUE)
                     
                     # last line inside the {} should be the output you want
                     return(theseHill)
                   })

# Summarizing this raw simulation, mean and SD are good options
neut_hillMeans <- apply(hillz, 1, mean)
neut_hillSDs <- apply(hillz, 1, sd)

## Comp sim hill 
hillz_2 <- replicate(500, # number of times to replicate
                   { # put all the code you need to replicate inside the { }
                           oneComp <- compSim(J = J, Sm = Sm, Jm = 10000, 
                                              nu = nu,m = m, alpha = a, niter = niter) 
                           
                           oneFinalSAD <- getAbund(oneComp, tstep = nrow(oneComp))
                           
                           # hill numbers
                           theseHill <- renyi(oneFinalSAD$abund, scales = 0:2, 
                                              hill = TRUE)
                           
                           # last line inside the {} should be the output you want
                           return(theseHill)
                   })


# Summarizing this raw simulation, mean and SD are good options
comp_hillMeans <- apply(hillz_2, 1, mean)
comp_hillSDs <- apply(hillz_2, 1, sd)


##Invasion sims

# Pre
hillz_4 <- replicate(500, # number of times to replicate
                     { # put all the code you need to replicate inside the { }
                             metaSAD <- pika::rlseries(S, 0.001) 
                             untbPre <- untbSim(J = J, nu = nu, m = m, niter = niter,
                                                initMeta = metaSAD) # pre invasion
                             
                             newMetaSAD <- metaSAD # modifying meta community
                             irare <- which.min(newMetaSAD)[1] # index of a rare species
                             newMetaSAD[irare] <- 2 * max(newMetaSAD) # amplifying rare x2
                             
                             untbPost <- untbSim(J = J, nu = nu, m = m, niter = niter, 
                                                 initLocal = untbPre[nrow(untbPre), ], 
                                                 initMeta = newMetaSAD) # post invasion
                             
                             abundPre <- getAbund(untbPre) # combining sims
                             abundPost <- getAbund(untbPost)
                             abundPost <- abundPost[abundPost$tstep != 0, ] # removing overlap
                             abundPost$tstep <- abundPost$tstep + max(abundPre$tstep) # timesteps post after pre
                             abundAll <- rbind(abundPre, abundPost) # combining
                             
                             oneFinalSAD <- abundPre$abund[abundPre$tstep == max(abundPre$tstep)]
                             
                             
                             # hill numbers
                             theseHill <- renyi(oneFinalSAD, scales = 0:2, 
                                                hill = TRUE)
                             
                             # last line inside the {} should be the output you want
                             return(theseHill)
                             
                     })

# Summarizing this raw simulation, mean and SD are good options
preinvas_hillMeans <- apply(hillz_4, 1, mean)
preinvas_hillSDs <- apply(hillz_4, 1, sd)

# Short-term post
hillz_5 <- replicate(500, # number of times to replicate
                     { # put all the code you need to replicate inside the { }
                             metaSAD <- pika::rlseries(S, 0.001) 
                             untbPre <- untbSim(J = J, nu = nu, m = m, niter = niter,
                                                initMeta = metaSAD) # pre invasion
                             
                             newMetaSAD <- metaSAD # modifying meta community
                             irare <- which.min(newMetaSAD)[1] # index of a rare species
                             newMetaSAD[irare] <- 2 * max(newMetaSAD) # amplifying rare x2
                             
                             untbPost <- untbSim(J = J, nu = nu, m = m, niter = niter, 
                                                 initLocal = untbPre[nrow(untbPre), ], 
                                                 initMeta = newMetaSAD) # post invasion
                             
                             abundPre <- getAbund(untbPre) # combining sims
                             abundPost <- getAbund(untbPost)
                             abundPost <- abundPost[abundPost$tstep != 0, ] # removing overlap
                             abundPost$tstep <- abundPost$tstep + max(abundPre$tstep) # timesteps post after pre
                             abundAll <- rbind(abundPre, abundPost) # combining
                             
                             oneFinalSAD <- abundPost$abund[abundPost$tstep == 13000]
                             
                             
                             # hill numbers
                             theseHill <- renyi(oneFinalSAD, scales = 0:2, 
                                                hill = TRUE)
                             
                             # last line inside the {} should be the output you want
                             return(theseHill)
                             
                     })
# Summarizing this raw simulation, mean and SD are good options
postinvas_hillMeans <- apply(hillz_5, 1, mean)
postinvas_hillSDs <- apply(hillz_5, 1, sd)

# Long-Term Post
hillz_6 <- replicate(500, # number of times to replicate
                     { # put all the code you need to replicate inside the { }
                             metaSAD <- pika::rlseries(S, 0.001) 
                             untbPre <- untbSim(J = J, nu = nu, m = m, niter = niter,
                                                initMeta = metaSAD) # pre invasion
                             
                             newMetaSAD <- metaSAD # modifying meta community
                             irare <- which.min(newMetaSAD)[1] # index of a rare species
                             newMetaSAD[irare] <- 2 * max(newMetaSAD) # amplifying rare x2
                             
                             untbPost <- untbSim(J = J, nu = nu, m = m, niter = niter, 
                                                 initLocal = untbPre[nrow(untbPre), ], 
                                                 initMeta = newMetaSAD) # post invasion
                             
                             abundPre <- getAbund(untbPre) # combining sims
                             abundPost <- getAbund(untbPost)
                             abundPost <- abundPost[abundPost$tstep != 0, ] # removing overlap
                             abundPost$tstep <- abundPost$tstep + max(abundPre$tstep) # timesteps post after pre
                             abundAll <- rbind(abundPre, abundPost) # combining
                             
                             oneFinalSAD <- abundPost$abund[abundPost$tstep == max(abundPost$tstep)]
                             
                             
                             # hill numbers
                             theseHill <- renyi(oneFinalSAD, scales = 0:2, 
                                                hill = TRUE)
                             
                             # last line inside the {} should be the output you want
                             return(theseHill)
                             
                     })

# Summarizing this raw simulation, mean and SD are good options
longinvas_hillMeans <- apply(hillz_6, 1, mean)
longinvas_hillSDs <- apply(hillz_6, 1, sd)

```

When evaluating the influence of invasion on diversity indices, the output is different is different for each simulation run. Because of this variability, there is no absolute value for species richness for any of the models. However, species richness across models is equivalent when taking the variability into consideration. Generally, species richness ranges from 22-29 across all models with the variation ranging from 1.5 to 4. Therefore, there is no significant difference in predicted species richness across all models. This trend carries across both the Shannon's and Inverse Simpson's indices for both the output and the variability. 



```{r, echo = FALSE, results='asis'}
## Compiling
hills_all <- rbind(neut_hillMeans,neut_hillSDs,
             comp_hillMeans, comp_hillSDs,
             preinvas_hillMeans, preinvas_hillSDs,
             postinvas_hillMeans, postinvas_hillSDs,
             longinvas_hillMeans,longinvas_hillSDs)

hills_all <-round(hills_all, digits = 2)

hills_all <- as.data.frame(hills_all)

hills_all <- hills_all %>% 
        rename( "Species Richness" = "0",
                "Shannons Index" = "1",
                "Inverse Simpsons" = "2")

row.names(hills_all) <- c("Neutral Model","SD Neut","Competitive Model","SD Comp","Pre-Invasion Model", "SD Invas1",
              "Short-Term Post Invasion","SD Invas2","Long-Term Post Invasion","SD Invas3")

kable(hills_all, caption = "Mean Biodiversity Indices")
```



**Challenges and Conclusions**

Historically, these models have been used to understand patterns of diversity across scales. We expected to see consistently higher values for all diversity metrics using the competitive model due to the understanding that competitive models take into consideration the impact of niche differentiation on diversity (Levine & HilleRisLambers 2009). We expected that invaders would have a negative impact on biodiversity because they might influence populations via competition for space or resource and increase predation interactions in the system (Tan & Beal 2015).

Applying neutral models to marine systems can be slightly more complex than applying the same model to terrestrial ecosystems. For one, dispersal limitation may play a more prominent role in ocean systems where both environmental factors (e.g. ocean currents) and marine life stages (e.g. planktonic larval stages) can influence community structure and biodiversity (Srisvastava and Pavel, 2013). In fact, studies have refuted the neutral model, positing that stochastic spatial-temporal environmental factors play a more important role in driving biodiversity patterns in marine systems (Connolly et al. 2014). For example, Dorenelas et al. (2006) found that coral reef biodiversity better matched with the classical log normal distribution than the neutral model. Further exploration into neutral models is needed as they have shown to be an effective way to model disturbance (Maa 2014). 

The usefulness of these models for understanding the impacts rare species introductions remains to be seen. Either the model didn't do a good job of evaluating system responses OR, under the neutral model the introduction of rare species does not influence biodiversity patterns. 

*Things to Consider*
Here are some considerations that may be confounding factors in using these models:
- Harvesting: One topic that should be considered in the future is the impact of harvesting on local and community biodiversity and the applications of these models to those processes. 
- Aquaculture practices: Includes the genetic manipulation of organisms, the introduction of rare species into systems, and the expansion of potential habitat via gear (Diana 2009).




**References**

Alzate, A., Janzen, T., Bonte, D., Rosindell, J., & Etienne, R. S. (2019). A simple spatially explicit neutral model explains the range size distribution of reef fishes. Global Ecology and Biogeography, 28(7), 875–890. https://doi.org/10.1111/geb.12899

Britsch, M. L. (2021). Marine Aquaculture in Maine: Understanding Diverse Perspectives and Interactions at Multiple Scales. University of Maine.

Delago, D. F. (2021). Investigating Larval Spillover From Oyster Aquaculture Through Geospatial Habitat Suitability Index Modeling: A Damariscotta River Estuary Case Study. University of New England.

Diana, J. S. (2009). Aquaculture production and biodiversity conservation. Bioscience, 59(1), 27-38.

Dolan, J. R., Ritchie, M. E., & Ras, J. (2007). The “neutral” community structure of planktonic herbivores, tintinnid ciliates of the microzooplankton, across the SE Tropical Pacific Ocean. Biogeosciences, 4(3), 297–310. https://doi.org/10.5194/bg-4-297-2007

Levine, J. M., & HilleRisLambers, J. (2009). The importance of niches for the maintenance of species diversity. Nature, 461(7261), 254-257.

Maine DMR. (2022). Aquaculture Map: Maine Department of Marine Resources. https://www.maine.gov/dmr/aquaculture/leases/aquaculturemap.html

Risley, S. (2022). Linking Local Knowledge & Community Science in Support of Coastal Marine Stewardship (3564) [The University of Maine]. Electronic Theses and Dissertations. https://digitalcommons.library.umaine.edu/etd/3564

Rosindell, J., Hubbell, S. P., & Etienne, R. S. (2011). The Unified Neutral Theory of Biodiversity and Biogeography at Age Ten. Trends in Ecology & Evolution, 26(7), 340–348. https://doi.org/10.1016/j.tree.2011.03.024

Shannon, C.E. & Weaver, W. (1949) The Mathematical Theory of
Communication. University of Illinois Press, Urbana.

Srivastava, D., & Kratina, P. (2013). Is dispersal limitation more prevalent in the ocean? Oikos, 122, 298–300.

Tan, E. B. P., & Beal, B. F. (2015). Interactions between the invasive European green crab, Carcinus maenas (L.), and juveniles of the soft-shell clam, Mya arenaria L., in eastern Maine, USA. Journal of Experimental Marine Biology and Ecology, 462, 62–73. https://doi.org/10.1016/j.jembe.2014.10.021

Whittaker, R. H. 1972. Evolution and measurement of species diversity. Taxon 21:213–251.

Wootton, J. T. (2005). Field parameterization and experimental test of the neutral theory of biodiversity—Nature. Nature, 433(7023), 309–312. https://doi.org/10.1038/nature03211







